---
title: "Drawing a 3D Cube in ggplot"
author: "mikefc"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
suppressPackageStartupMessages({
  library(ggplot2)
  library(magrittr)

  library(threed)
})


knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



```{r}
library(ggplot2)

library(threed)
```


Introduction
-------------------------------------------------------------------------------

Prepare an object for plotting
-------------------------------------------------------------------------------

* Create an object (here the standard 2x2x2 cube is being used)
* Define where camera is located, and where it is looking
* Transform the object into camera space
* Perspective transform the data


```{r}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# The `threed` package has some builtin objects in `threed::mesh3dobj`
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
obj <- threed::mesh3dobj$cube

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Define camera 'lookat' matrix i.e. camera-to-world transform
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
camera_to_world <- threed::look_at_matrix(eye = c(1.5, 1.75, 4), at = c(0, 0, 0))

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Transform the object into camera space and do perspective projection
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
obj <- obj %>%
  transform_by(invert_matrix(camera_to_world)) %>%
  perspective_projection()
```


```{r}
as.data.frame(obj) %>% knitr::kable()
```



Plot the points for the vertices of the object
-------------------------------------------------------------------------------

* `threed` defines a `fortify.mesh3d()` function.
* If a `mesh3d` object is given as the data for a`ggplot2` call, `ggplot2` will 
  automatically use `fortify()` to convert into a data.frame.
* i.e. because `threed` defines `fortify.mesh3d()`, we can call `ggplot2` directly with a `mesh3d` object.

```{r}
ggplot(obj, aes(x, y)) +
  geom_point() +
  theme_void() +
  theme(legend.position = 'none') +
  coord_equal()
```


Plot the outline of each polygon
-------------------------------------------------------------------------------

* Each element has a unique `element_id`, and this is used as the `group` aesthetic 
  to inform ggplot that it should draw one polygon for each element.
* Set `fill = NA, colour = 'black'` to draw only the borders of each polygon.

```{r}
ggplot(obj, aes(x, y, group = element_id)) +
  geom_polygon(fill = NA, colour = 'black', size = 0.2) +
  theme_void() +
  theme(legend.position = 'none') +
  coord_equal()
```


Dotted rendering of hidden lines
-------------------------------------------------------------------------------

* To draw *hidden* elements in a different way, use the `hidden` variable.
* `hidden` is a boolean variable indicating if a triangle or quad element is facing
  away from the camera.
* Here a different `linetype` is used for hidden elements 

```{r}
ggplot(obj, aes(x, y, group = element_id)) +
  geom_polygon(fill = NA, colour='black', aes(linetype = hidden,  size = hidden)) +
  scale_linetype_manual(values = c('TRUE' = "dotted", 'FALSE' = 'solid')) +
  scale_size_manual(values = c('TRUE' = 0.2, 'FALSE' = 0.5)) +
  theme_void() +
  theme(legend.position = 'none') +
  coord_equal()
```


Hidden line removal
-------------------------------------------------------------------------------

* Here a zero-width `linetype` is used for hidden elements to hide them.

```{r}
ggplot(obj, aes(x, y, group = element_id)) +
  geom_polygon(fill = NA, colour = 'black', aes(size = hidden)) +
  scale_size_manual(values = c('TRUE' = 0, 'FALSE' = 0.5)) +
  theme_void() +
  theme(legend.position = 'none') +
  coord_equal()
```


Naive Filled Polygons
-------------------------------------------------------------------------------

* When filling polygons, `ggplot2` will draw them in the order of the `group` variable.
* But (as is usually the case), elements drawn in this order will overlap incorrectly 
  with other elements, and the result will look weird - Almost Escher-esque!

```{r}
ggplot(obj, aes(x, y, group = element_id)) +
  geom_polygon(fill = 'lightblue', colour = 'black', size = 0.2) +
  theme_void() +
  theme(legend.position = 'none') +
  coord_equal()
```


Filled Polygons - (1) Drop hidden elements so they never get drawn
-------------------------------------------------------------------------------

* **First method** for drawing filled polygons:
    * **drop hidden elements manually**
    * Do this by explicitly converting the `mesh3d` object to a data.frame, and then filtering
      out the hidden elements.

```{r}
obj_df <- as.data.frame(obj) 
obj_df <- subset(obj_df, !obj_df$hidden)

ggplot(obj_df, aes(x, y, group = element_id)) +
  geom_polygon(fill = 'lightblue', colour = 'black', size = 0.2) +
  theme_void() +
  theme(legend.position = 'none') +
  coord_equal()
```




Filled Polygons - (2) Assign hidden elements a fill of `NA` so they get drawn invisibly
-------------------------------------------------------------------------------

* **Second method** for drawing filled polygons: 
    * **draw hidden elements invisibly**
    * Do this by explicitly converting the `mesh3d` object to a data.frame, and
      then manually setting the variables mapped to `fill` and `colour` to 
      be `NA` for hidden elements

```{r}
obj_df <- as.data.frame(obj) %>%
  transform(
    shade      = ifelse(hidden, NA, 'lightblue'),
    linecolour = ifelse(hidden, NA, 'black')
  )

ggplot(obj_df, aes(x, y, group = element_id)) +
  geom_polygon(aes(fill = I(shade), colour = I(linecolour)), size = 0.2) +
  theme_void() +
  theme(legend.position = 'none') +
  coord_equal() 
```



Filled Polygons - (3) Use the `zorder` variable to control draw order
-------------------------------------------------------------------------------


* **Third method** for drawing filled polygons: 
    * **Draw the elements from furtherest to nearest**
    * Exploit the fact that elements are drawn in the order of the `group` variable.
    * When converting a `mesh3d` to a data.frame, a `zorder` variable is created starting at
      `1` for the furtherest element, up to `n` for the closest element.
    * i.e. change the `group` variable from `element_id` to `zorder`
    
```{r}
ggplot(obj, aes(x, y, group = zorder)) +
  geom_polygon(fill = 'lightblue', colour='black', size = 0.2) +
  theme_void() +
  theme(legend.position = 'none') +
  coord_equal() 
```



Fake-shaded polygon
-------------------------------------------------------------------------------

* The normal to each face is included in the data.frame representation as `fnx, fny, fnz`
* By calculating dot products to a light source positioned in the scene, the fraction 
  illumination could be calculated for each element.
* Here, the shading is being completely faked by using the sum `fny + fnz` to shade the polygons.

```{r}
ggplot(obj, aes(x, y, group = zorder)) +
  geom_polygon(aes(fill = fny + fnz), colour = 'black', size = 0.2) +
  theme_void() +
  theme(legend.position = 'none') +
  coord_equal()

```


