---
title: "threed + mesh3d objects"
author: "mikefc"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


mesh3d adaptations
------------------------------------------------------------------------------

The `threed` package adapts/extends the `mesh3d` format to include the following:

* Meshes now support `primitivetype` of `point` and `line` 
  (in addtion to the standard `triangle` and `quad`)
* Meshes with line segments will contain `il`. This is a `2 x n` matrix giving
  the indices of the vertex of each end of the line segment.
* Meshes with point data will contain `ip`.  This is a `1 x n` matrix giving
  the index of the vertex of the point.
* A general `properties` data.frame can be included in the object. The number of
  rows in this data.frame must equal the total number of elements in the object.
* A `transform_matrix` is a 4x4 transformation matrix. If missing or NULL, then it
  is treated as the identity matrix.
  



Transforming `mesh3d` objects
------------------------------------------------------------------------------

* The supported object transformations are:
    * `scale_by(obj, c(x, y, z))`
    * `translate_by(obj, c(x, y, z))`
    * `rotate_by(obj, angle, c(x, y, z))`
    * `perspective_projection(obj, w, h, n, f)
* When applying a transform to an object, the object vertices are not immediately transformed.
* Instead, a `mesh3d` object now includes an extra `transform_matrix` element which
  keeps track of the transforms applied to this object.
* To actually update the vertex positions, call `actualize_transformation` on the object.

```{r eval=FALSE}
# A 2x2x2 cube centred at the origin
threed::mesh3dobj$cube 

# Transforming the cube only updates the `transform_matrix` of the object
threed::mesh3dobj$cube %>%
  threed::translate_by(c(x=0, y=1, z=1))

# Actualizing the transformation updates the vertices
threed::mesh3dobj$cube %>%
  threed::translate_by(c(x=0, y=1, z=1)) %>%
  actualize_transformation()
```




`as.data.frame.mesh3d`
------------------------------------------------------------------------------

The `as.data.frame.mesh3d` function will convert a `mesh3d` object into a
very verbose data.frame.

This function first calls `threed::actualize_transformation()` to apply any 
outstanding transformations to the vertices.

The resulting data.frame can then be used to plot the 3d points, lines and polygons 
in `ggplot2`.  It has the following columns by default:

* `element_id` - element index
* `element_type` - an integer from 1 to 4 indicating whether the element is a point, line, triangle or quad
* `vorder` - vertex ordering within element
* `x,y,z` - coorindates of vertex
* `vertex` - global index of this vertex
* `vnx,vny,vnz` - unit vector in direction of vertex normal
* `fnx,fny,fnz` - unit vector in direction of face normal
* `fcx,fcy,fcz` - coordinates of centroid of element
* `zorder` - zorder is used to control ggplot draw order. For polygons it is based upon fcz. For lines it is the smaller of the z coordinates. For points, it is the z coordinate of the point
* `zorder_var` - The value used to determine zorder
* `hidden` - by assuming the camera faces along the negative z-axis, any triangle or quad face which has a face normal which also points along the negative z-axis faces away from the camera and is hidden. Boolean value.


```{r}
obj <- threed::mesh3dobj$cube

as.data.frame(obj)
```








